---
title: "Are we more civilized today?"
author: "Written by group members: Benjamin, Lasse & Simon (Group 13)"
date: "created 4 October 2021, update:`r format(Sys.Date(),'%e of %B, %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE)
```
This exercise is based on the dataset provided by [OurWorldInData project](https://ourworldindata.org/homicides) based at the Oxford University.


# The long-term trend in Homicides in Western Europe

Understanding how homicide rates have changed prior to the modern era requires the help of historians and archivists. Manuel Eisner, a criminology professor at the University of Cambridge, and his colleagues published the Historical Violence Database \href{https://cjrc.osu.edu/research/interdisciplinary/hvd}: a compilation of data on long-term trends in homicide rates, in addition to qualitative information such as the cause of death, perpetrator and victim. This database is limited to countries with relatively complete historical records on violence and crime – mainly Western Europe and the US.

Starting in the second half of the nineteenth century, these European regions have consistent police records of those accused of murder or manslaughter and annual counts of homicide victims. To go back further in time, reaching as far back as the thirteenth century, Eisner collected estimates (from historical records of coroner reports, court trials, and the police) of homicide rates made in over ninety publications by scholars.

Homicide rates – measured as the number of homicides per 100,000 individuals – up to 1990 are sourced from Eisner’s (2003) publication and the Historical Violence Database. 

Are homicide rates in Europe today lower or higher than in the past? Using the provided dataset, display and describe the long-run homicide rates for the five European regions: Italy, England, Germany, Netherlands and Scandinavia.

```{r library}
library(tidyverse)
```

Code and text written by: Simon

This code loads the tidyverse package and makes certain bits of code available like "read_csv2", which will be used later in the part of the assignment that works on the reigning periods of danish monarchs.




# Load the available data from ourworldindata.org

You should always interrogate the source of your data. Who compiled it, from where, what is missing, how representative the data are? Check the data/Metadata.txt to learn about the data provenance.
```{r load-data}
Western_Europe <- read_csv("data/homicide-rates-across-western-europe.csv")
```

Code and text written by: Benjamin

This takes the csv-file "homicide-rates-across-western-europe" and attaches it to "Western_Europe" so when that item is actiavated it will load the csv-file.






# Inspect the data

How clean and analysis-ready is the dataset? Do you understand what the column names represent? What is the difference between rate and homicide number?
```{r inspect}
head(Western_Europe)
```

Code and text written by: Benjamin

By writing head(Western_Europe) we only get the upper portion of the dataset, which is called the "head". 
The first column "Entity" represents the country from which that data is from. The third column "Year" represents the year the data is from. 
The fourth column "Homicide rate in Europe over long-term (per 100,000) (homicides per 100,000 people)" represents how many people were killed per 100 thousand in the different countries per year.
The difference between rate and homicide number in the year 1300 is that the rate is  that 23 people died per 100 thousand, while the homicide number is the total amount of homicides for the country in the year 1300.






Ok, the data look good except for the column `Homicide rate in Europe over long-term (per 100,000)` which is not very easy to work with.

- Use the `names()` function and assignment key to relabel this column to `homicides_per_100k` 

```{r relabel column}
names(Western_Europe)[names(Western_Europe) == "Homicide rate in Europe over long-term (per 100,000) (homicides per 100,000 people)"] <- "homicides_per_100k"
```

Code and text written by: Benjamin

This code selects a column and changes the columns name into the name requested by the task.
The assignment operator "<-" is replacing "Homicide rate in Europe over long-term (per 100,000) (homicides per 100,000 people)" with "homicides_per_100k".
The square brackets selects the item "Homicide rate in Europe over long-term (per 100,000) (homicides per 100,000 people)" inside of the group "names(Western_Europe)", whereafter the assignment operator "<-" replaces the chosen column name in "names(Western_Europe)" with the requested shorter name "homicides_per_100k".




Now, that you have looked at what the data looks like and what it represents, and streamlined it, let's see what big picture it contains.


# Let's see what the long-term trend is in homicides


- use `ggplot()` function and remember the`+` at the end of the line
- chose a meaningful `geom_......()`  for geometry (hint: points are not great)
- load `Year` on the `x` axis and `homicides_per_100k` column in y axis
- to color individual country entries consistently, assign the country column to the argument `color`.
- provide meaningful title and axis labels
- remember to change the `eval` flag so that the code chunk renders when knitted 

```{r plot-homicides-territories-together, eval = TRUE}
ggplot(data = Western_Europe,aes(x = Year, y = homicides_per_100k, color = Entity)) + 
  geom_line()
```

Code and text written by: Simon

This code creates a graph from the dataset (Western_Europe).
"ggplot(data = Western_Europe" means that the graph will be based on the data (Western_Europe), which is about homicides in countries across several years.
"x = Year, y = homicides_per_100k" is telling the graph what variables you want on both the x and y axis.
And then to make the graph more digestible we add color to the lines with the "color = Entity" code, while "geom_line()" makes it into a graph that uses lines specifically.







Alright, the homicide rates should all be descending over time. What a comfort. But the viz is not super clear. Let's check the rates for individual countries.

# Uncouple the homicides of individual countries for easier view

You can visualize each country's trend separately by adding an extra argument to the ggplot, the `facet_wrap()` and feeding it the country column. If in doubt, check your ggplot tutorial and your country column name for exact usage.

- reuse the ggplot from the chunk above
- insert `facet_wrap()` after the specification of geometry to split countries in separate charts
- change the facet "layout" to two columns and three rows so that the trends are easier to see in horizontal layout. 


```{r Territories-Separate, eval = TRUE}
ggplot(data = Western_Europe, aes(x = Year, y = homicides_per_100k, color = Entity)) + 
  geom_line() +
  facet_wrap("Entity") +
  theme(legend.position = "bottom") +
  labs(color = "Country")
  

```

Code and text written by: Simon

This code in essence does the same as the previous task, where the difference is that both the legend has been placed on the bottom by the "theme(legend.position = "bottom")" code, the different countries/territories have been split up by the "facet_wrap("Entity")" code, 
and the "labs(color = "Entity")" code renames the legend from "Entity" to "Country".





# Compare the trends in homicide with the pattern of reign duration among Danish rulers through time.

- Load your Danish king dataset. Hopefully it is tidy and your years and duration of reign are all numeric.
- You need to have a consistent way of plotting the rulers' reign on the x axis, so I recommend you create a midyear column by calculating the middle of each monarch's rule (Hint: `midyear` = `endyear` - (`endyear`-`startyear`)/2)
- Start a ggplot plotting midyear on x axis and duration on y axis
- Try `geom_smooth()` for geometry 
- Provide meaningful labels and a title
- How would you characterize the trend compared to the homicides above?

```{r Kings-ggplot, eval = TRUE}
kongerækken <- read_csv2("data/Uge10Kongeraekke.csv")
kongerækken <- kongerækken %>% 
  mutate(midyear = Regering_slut-(Regering_slut-Regering_start)/2) %>% 
  mutate(reign = Regering_slut - Regering_start)

ggplot(data = kongerækken,aes(x = midyear, y = reign)) + 
  geom_smooth() +
  labs(title = "Regeringsperioder over tid")
```

Code and text written by: Lasse 

Here the "read_csv2" code is used as the file we created in week 10, was semicolon separated, and "read_csv2" reads semicolon separated files.
The code used with mutate was to create the item midyear, and the item reign.
Which were then both used to create the ggplot's x and y axis.
Those being "(x = midyear, y = reign)".
The code "labs(title = "Regeringsperioder over tid")" then creates the title of the graph, which is "Regeringsperioder over tid".
The code "geom_smooth()" displays the statistical uncertainty and the mean of the reigns.

And to finally compare this graph with homicide it would seem, that since the monarchs are allowed to stay on the throne for longer, and the homicide rate is going down, that we may be more civilized. There are no violent uprisings to claim the throne, and most monarchs only hold symbolic power. This coupled with people getting murdered less, could show a tendency towards less violent actions in society.


# Final tasks.

1) Plot: In the facetted plot above, move the legend from the current position on the side to below the facets, and label it "Country" instead of "Entity".

Code written by: Simon

2) Rmarkdown: 

  - edit the author of the document, and convert 'Final Tasks' into heading #2 (like the other headings) 
  - add a `floating table of contents` to your Rmarkdown document, 
  - provide informative `chunk-names` and edit flags in your R chunks, and
  - automatically generate a `timestamp` to show when the document was last updated. (Hint: check the Rmarkdown episode in our Data Carpentry tutorial)

Code and text written by: Benjamin

The author of the document has been edited to be the group members of group 13.
Both the flags, and the chunk names of RMD code chunks have been edited.
Timestamp has been added for when it is updated.
The floating table of contents is available in the HTML

3) Question: In <250 words articulate your answer on the basis of the data visualisations to the following question: are we more civilized today?

Text written by: Lasse

The way that society works in the present has more regulations, and has more safety nets, both to keep people safe, like police, and social safety nets like welfare given by the state so people don't get pushed to the mental brink and commit crimes.
Homicides still happen, but there's less of them, my belief is that this is because less people have to get into situations where they're forced to take somebody elses life, like a home invasion, or a robbery gone wrong.
This however has nothing to do with the longer reigns of monarchs, as most countries now are democracies, and most monarchs only have symbolic power, monarchs being in power longer has nothing to do with less murders, both are a symptom of a perhaps more civilized society.

Written by group members: Benjamin, Lasse & Simon (Group 13)

See Github link: 

Lasse: https://github.com/Digital-Methods-HASS/Au753705_Lasse_Walmar/tree/main/Week_12_Assignment

Benjamin: https://github.com/Digital-Methods-HASS/202108995_Rosenstr-m_Benjamin/tree/main/Week%2012 

Simon: https://github.com/Digital-Methods-HASS/690090_Pedersen_Simon/tree/main/Week%2012 
